Comparing R packages for calculation of persistent homology

### 2020-03-19 tdabench-paper/tdabench.Rmd
capitlization

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. The Dionysus (TDA), GUDHI (TDA), and Ripser C++ libraries (TDAstats) provide Rips complex functionality. GUDHI  (TDA package) also provides Alpha complex functionality. These software represent powerful analysis tools but are limited computationally. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. We benchmarked the all libraries on a High-Powered Cluster for runtime. We also compared Alpha and Rips for memory burden. Alpha complexes run significantly faster than all Rips complex libraries for all data sets for data dimension less than or equal to three. Alpha complexes were not computable for data dimension greater than 3. TDAstats had the fastest runtimes for all Rips complexes. We recommend performing persistent homology with GUDHI Alpha complexes for data dimension less than or equal to three. For higher dimension, we recommend TDAstats Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

\section{Introduction}

Topological data analysis (TDA) is a broad set of methodologies that characterize structural features of data sets inspired by topological principles.
Within the umbrella of TDA, persistent homology represents an algebraic approach to understanding the number, characteristics, and persistence of structural features in an n-dimensional point cloud.

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Rips and Alpha complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

@@

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Rips complexes in the \CRANpkg{TDAstats} and \CRANpkg{TDA} R packages.
We also explore the differences in computational performance between Rips and Alpha complexes in R to assess which datasets are most appropriate for each algorithm.

@@

Large data and feature dimensionality often restricts persistent homology calculations to small point clouds due to computational limits.
When calculating persistent homology on a high-dimensional point cloud, as Vietoris-Rips feature dimension increases, there is a corresponding increase in runtime (Figure \ref{fig:box}).
Dionysus is clearly outmatched by GUDHI and Ripser as feature dimension increases, with the difference being clearest for larger point clouds; by feature dimension 5, Ripser outpaces GUDHI as well (Figure \ref{fig:box}).
Due to the limited number of points, it is unclear whether runtime for each library grows polynomially or exponentially.
<!--RECHECK, should be enough data point to regress on-->

@@

Even with a constant feature dimension, the underlying data dimension could play a role in runtime of persistent homology calculation.
Figure \ref{fig:ann} compares handling of this issue by the Vietoris-Rips complex and the alpha complex.
Since GUDHI is the only library implementing functionality with an alpha complex, we compare its implementations of the Vietoris-Rips and alpha complices.
Due to computational limitations, an alpha complex could not be calculated for any point clouds with data dimension exceeding 3.
Two notable aspects of Figure \ref{fig:ann} stand out.
First, the alpha complex calculation clearly runs faster than the Vietoris-Rips complex calculation, a trend that becomes clearer as point cloud size increases.
Second, although the Vietoris-Rips complex calculation runtime appears to be independent of underlying data dimension, alpha complex calculation is dependent on it.
Figure \ref{fig:ann} shows a subtle difference between data dimensions 2 and 3 as point cloud size increases.
Although unconcerning for a data dimension up to 3, failure to run any alpha complex calculations with a data dimension of 4 could be cause for concern.

>>>>

---
title: Comparing R packages for calculation of persistent homology
author:
  - name: Eashwar V. Somasundaram
    affiliation: School of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44106, United States
    email: eashwar.somasundaram@case.edu
  - name: Shael E. Brown
    affiliation: Department of Quantitative Life Sciences
    address:
    - McGill University
    - Montreal, QC H3A 0G4, Canada
    email: shael.brown@mail.mcgill.ca
  - name: Adam Litzler
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email:  adli4611@colorado.edu
  - name: Jacob G. Scott
    affiliation: Department of Translational Hematology and Oncology Research
    address:
    - Cleveland Clinic Foundation
    - Cleveland, OH 44195, United States
    email: ScottJ10@ccf.org
  - name: Raoul R. Wadhwa
    affiliation: Cleveland Clinic Lerner College of Medicine
    address:
    - Case Western Reserve University
    - Cleveland, OH 44195, United States
    email: raoul.wadhwa@case.edu
abstract: >
  Several persistent homology software libraries have been implemented in R. The Dionysus (TDA), GUDHI (TDA), and Ripser C++ libraries (TDAstats) provide Rips complex functionality. GUDHI  (TDA package) also provides alpha complex functionality. These software represent powerful analysis tools but are limited computationally. These limitations and guidelines for which persistent homology libraries to use have not been specified in R. We benchmarked the all libraries on a High-Powered Cluster for runtime. We also compared alpha and Rips for memory burden. Alpha complexes run significantly faster than all Rips complex libraries for all data sets for data dimension less than or equal to three. Alpha complexes were not computable for data dimension greater than 3. TDAstats had the fastest runtimes for all Rips complexes. We recommend performing persistent homology with GUDHI alpha complexes for data dimension less than or equal to three. For higher dimension, we recommend TDAstats Rips Complex. 
output:
  rticles::rjournal_article:
    includes:
      in_header: preamble.tex
---

\section{Introduction}
Topological data analysis (TDA) is a broad set of methodologies that characterize structural features of data sets inspired by topological principles.
Within the umbrella of TDA, persistent homology represents an algebraic approach to understanding the number, characteristics, and persistence of structural features in an n-dimensional point cloud.

In the basic workflow of persistent homology, a series of simplicial complexes are generated on point clouds to characterize the topological features. 
Simplicial complexes are connected sets of points, lines, and n-dimensional shapes.
There are several methods to generate these complexes on point clouds. 
In this paper, we are using the Rips and alpha complex for our persistent homology calculations. 
In both methods, a series of data structures, called simplicial complexes, are built on top of a point cloud.
These complexes can be viewed as approximating real topologic relationships in the data.
The exact method of constructing these complexes is described in the Mathematics section. 
However, the key takeaway is that the simplicial complex attempts to reveal the underlying topology of the point cloud. 
These features can be described by dimension and "persistence." 
Long lasting (persistent) features that exist across the series of simplicial complexes intuitively represent "real" geometrical patterns as opposed to noise.

@@

Topological data analysis has had a broad range of usage from viral evolution to physical chemistry\cite{TDA-Viral,TDA-PChem}. 
Given that persistent homology can apply to any n-dimensional data set, it has wide ranging applicability. 
Its main limiting factor is computational for calculating high dimensional features in large multi-dimensional data sets. 
We believe lack of analysis of these higher dimensional features may under utilize persistent homology's analysis potential.
In this project, we explore different methodologies to find efficient options that could solve this issue. 
We specifically aim to characterize the computational performance (run time) and limits of persistent homology calculations between Rips complexes in the R packages TDAstats and TDA.
We also explore the differences in computational performance between Rips and alpha complexes in R to assess which datasets are most appropriate for each algorithm.

@@

<!--FOLLOWING SENTENCES SHOULD GO IN DISCUSSION
For an n-dimensional point cloud, up to n-1 feature dimensions can be calculated. 
In the field, most work has only included calculations up to 1 or 2 feature dimensions.
Perhaps this is due to difficulty of conceptualizing what high dimensional features represent. 
However, it might also be due to computational restrictions of calculating high dimensional features.-->
Large data and feature dimensionality often restricts persistent homology calculations to small point clouds due to computational limits.
When calculating persistent homology on a high-dimensional point cloud, as Vietoris-Rips feature dimension increases, there is a corresponding increase in runtime (Figure \ref{fig:box}).
Dionysus is clearly outmatched by GUDHI and Ripser as feature dimension increases, with the difference being clearest for larger point clouds; by feature dimension 5, Ripser outpaces GUDHI as well (Figure \ref{fig:box}).
Due to the limited number of points, it is unclear whether runtime for each library grows polynomially or exponentially.
<!--RECHECK, should be enough data point to regress on-->

@@

<!--SHOULD BE IN METHODS
The GUDHI library provided functionality to calculate persistent homology with Rips and Alpha complexes. 
Therefore, we only used the GUDHI library to compare performance between Rips and Alpha complexes.-->

Even with a constant feature dimension, the underlying data dimension could play a role in runtime of persistent homology calculation.
Figure \ref{fig:ann} compares handling of this issue by the Vietoris-Rips complex and the alpha complex.
Since GUDHI is the only library implementing functionality with an alpha complex, we compare its implementations of the Vietoris-Rips and alpha complices.
Due to computational limitations, an alpha complex could not be calculated for any point clouds with data dimension exceeding 3.
Two notable aspects of Figure \ref{fig:ann} stand out.
First, the alpha complex calculation clearly runs faster than the Vietoris-Rips complex calculation, a trend that becomes clearer as point cloud size increases.
Second, although the Vietoris-Rips complex calculation runtime appears to be independent of underlying data dimension, alpha complex calculation is dependent on it.
Figure \ref{fig:ann} shows a subtle difference between data dimensions 2 and 3 as point cloud size increases.
Although unconcerning for a data dimension up to 3, failure to run any alpha complex calculations with a data dimension of 4 could be cause for concern.
<!--SHOULD BE IN DISCUSSION
This is understandable as Alpha complex theoretical complexity increases exponentially for increasing point cloud dimension.
Changes in computational performance from changing point cloud dimension is one major difference between calculation of Rips and Alpha complexes.-->
